---
title: "Homework 2"
subtitle: "BIOS 684"
author: "Jeff Oliver"
date: "15 September 2017"
output: 
  pdf_document:
  latex_engine: xelatex
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

<!--
DUE: Wednesday 20 September 2017

Instructions:
1. For each question requiring data analysis, support your conclusions by including the
relevant SAS output in your answer.
2. Include your SAS program as an Appendix to your solutions (PLEASE don’t include
the output from your program!).

Joint Modeling of Mean and Covariance in the Dental Growth Study
In a study of dental growth, measurements of the distance (mm) from the center of the
pituitary gland to the pteryomaxillary fissure were obtained on 11 girls and 16 boys at
ages 8, 10, 12 and 14 years (Potthoff and Roy, 1964).

The data are in the file “dental.txt” on the D2L dataset folder. Each row of the data
set contains the following six variables: subject ID, gender (“F” or “M”), and the
measurements at ages 8, 10, 12 and 14 years, respectively.

For all analyses in this homework, subset the dataset to contain only the
measurements for the girls.
-->

```{r echo = FALSE}
# Read in data
dental.data <- read.csv(file = "data/dental.csv",
                        header = FALSE)
# Add column names
colnames(dental.data) <- c("ID", "gender", "m.8", "m.10", "m.12", "m.14")

# Subset data
dental.data <- dental.data[dental.data$gender == "F", ]
```


## PART A: Descriptive Analyses
### 1. Plot the observed trajectories of distance for the 11 girls (all on one plot). Briefly, describe the important patterns that are apparent in the plot. (hint: use proc gplot)

The plot below shows pairwise comparisons between measurements of the same subject at different time points (below diagonal) and the correlation coefficient (r^2^) for the relationship (above diagonal):
```{r echo = FALSE}
# Function to print correlation coefficient; copied shamelessly from pairs documentation
panel.cor <- function(x, y, digits = 2, prefix = "", cex.cor, ...) {
    usr <- par("usr"); on.exit(par(usr))
    par(usr = c(0, 1, 0, 1))
    r <- abs(cor(x, y))
    txt <- format(c(r, 0.123456789), digits = digits)[1]
    txt <- paste0(prefix, txt)
    if(missing(cex.cor)) cex.cor <- 0.8/strwidth(txt)
    text(0.5, 0.5, txt, cex = cex.cor * r)
}

# Create scatterplot of matrices to inspect data
pairs(dental.data[, 3:6], 
      labels = c("Week 8", "Week 10", "Week 12", "Week 14"),
      upper.panel = panel.cor)
```

There is very high within-subject correlation among time samples. Interestingly, the within-subject correlation does not appear to decrease much with time (e.g. the correlation coefficient between week 8 and week 10 is effectively the same as between week 8 and week 14).

### 2. Obtain descriptive statistics for the distances in girls including means and standard deviations at each measurement occasion. Plot the means against time and comment on the shape of their trajectory. Also, comment on how the standard deviations are changing with time.

### 3. Obtain the variance-covariance and correlation matrices for the repeated measurements over time. Briefly, describe the important patterns in the correlations.

### 4. Identify one unexpected feature of this correlation structure generated in question 3. Provide a possible reason for this unexpected feature.

## PART B: Repeated Measures Models
### 5. Use PROC MIXED to fit a repeated measures model in which there is separate parameter for the mean distance at each of the four ages [HINT: Use the NOINT option on the MODEL statement]. Use an unstructured variance-covariance structure. For estimating parameters, fit the model using (i) the METHOD=ML option and (ii) the METHOD=REML option on the PROC MIXED statement.

#### a. Does either method of estimation give the same estimates of the variances and covariances as in your answer to question 3? If so, which one or ones?

#### b. Compare the variances and covariances for the two methods of estimation (i.e. ML versus REML). If they are not the same, comment on why they are different and state which method should be preferred and why?

**Check & revise language**: Different because ML is a biased estimator of the variance-covariance matrix ($\sigma$)because it is based on estimated beta hats; in contrast, REML <description of what REML is doing to create an unbiased estimator of the variance-covariance matrix> is an unbiased esimator of $\sigma$. Describe when to use which one.

#### c. Compare the estimated betas in the regression model to the sample means calculated in question 2. Provide a reason for any relationship you notice between these two sets of estimates.

### 6. Using METHOD=ML and an unstructured variance-covariance structure, fit a model which assumes that the mean distance changes linearly with time.

#### a. Why can this model be considered to be nested within the model fitted in question 5?

#### b. By undertaking a likelihood ratio test, evaluate the goodness of fit of this model compared with the model fitted (with METHOD=ML) in question 5. Which model do you prefer and why?

#### c. Why should you not use the REML log likelihood to compare models in this question?
Describe how REML creates non-nested models
