---
title: "Homework 8"
subtitle: "BIOS 684"
author: "Jeff Oliver"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output: 
  pdf_document:
  latex_engine: xelatex
---

```{r setup, include=FALSE, fig.height = 4}
knitr::opts_chunk$set(echo = FALSE, dev = "pdf")
library("dplyr")
library("ggplot2")
library("gee")
library("geepack")
```

***

<!--
Background:

The Skin Cancer Prevention Study, a randomized, double-blind, placebo-controlled clinical trial, was designed to test the effectiveness of beta-carotene in the prevention of non-melanoma skin cancer in high-risk subjects. A total of 1,683 subjects were randomized to either placebo or 50mg of beta-carotene per day and were followed for up to 5 years. Subjects were examined once per year and biopsied if a cancer was suspected to determine the number of new cancers per year.
The outcome variable, Y, is a count of the number of new skin cancers per year.

Selected data from the study are in the dataset called “skin.txt” on the course web site. Each row of the dataset contains the following 9 variables: ID, Center, Age, Skin, Gender, Exposure, Y, Treatment, Year. These variables take values as follows:
ID: Subject identifier number
Center: Identifier number for center of enrollment.
Age: Subject’s age in years at randomization
Skin: Skin type (1=burns; 0 otherwise) [evaluated at randomization and doesn’t change with time]
Gender: 1=male; 0=female
Exposure: Count of number of previous skin cancers [prior to randomization]
Y: Count of number of new skin cancers in the Year of follow-up
Treatment: 1=beta-carotene; 0=placebo
Year: Year of follow-up after starting randomized treatment

You may assume that the counts of new skin cancers, Y, are from exact one-year periods (so that no offset term is needed).
-->

```{r read-data}
# Read data
skin.data <- read.delim(file = "data/skin.txt", sep = " ", header = FALSE)

# Trim first & last columns (created by whitespace)
skin.data <- skin.data[, -c(1, ncol(skin.data))]

# Identify columns
colnames(skin.data) <- c("id", "center", "age", "skin", "gender", "prior.count", "new.count", "treatment", "year")

# Set factors as needed
factor.cols <- c("id", "center", "skin", "gender", "treatment")
for (c in factor.cols) {
  skin.data[, c] <- factor(skin.data[, c])
}
```

### 1. Your collaborator is interested in assessing the effect of treatment on the incidence of new skin cancers over time. As the statistician on the project, provide an analysis of the data that addresses this question. Please,
#### i) Provide a short table providing a descriptive summary of the mean count of new skin cancers by Treatment for each Year. Briefly comment on changes in incidence of new skin cancers by randomized treatment over time.

```{r descriptive-summary}
skin.description <- skin.data %>%
  group_by(treatment, year) %>%
  summarize(mean.new.count = mean(new.count),
            se.new.count = sd(new.count)/sqrt(length(new.count)))

skin.table <- data.frame(year = skin.description$year[1:5],
                         treat.0 = round(skin.description$mean.new.count[1:5], 3),
                         treat.1 = round(skin.description$mean.new.count[6:10], 3))

```

| Year | Placebo | Beta-carotene |
|:----:|:-------:|:-------------:|
| `r skin.table$year[1]` | `r skin.table$treat.0[1]` | `r skin.table$treat.1[1]` |
| `r skin.table$year[2]` | `r skin.table$treat.0[2]` | `r skin.table$treat.1[2]` |
| `r skin.table$year[3]` | `r skin.table$treat.0[3]` | `r skin.table$treat.1[3]` |
| `r skin.table$year[4]` | `r skin.table$treat.0[4]` | `r skin.table$treat.1[4]` |
| `r skin.table$year[5]` | `r skin.table$treat.0[5]` | `r skin.table$treat.1[5]` |

```{r descriptive-figure, fig.height = 3}
means.plot <- ggplot(data = skin.description, mapping = aes(x = year, y = mean.new.count, color = treatment, group = treatment)) +
  geom_errorbar(mapping = aes(x = year, ymin = (mean.new.count - se.new.count), ymax = (mean.new.count + se.new.count), width = 0.15)) +
  geom_point() +
  geom_line() +
  scale_color_manual(name = "Treatment", values = c("#000000", "#EC6E35"), labels = c("Placebo", "Beta-carotene")) +
  xlab(label = "Year") +
  ylab(label = "Mean new cancers")

print(means.plot)

```
Note error bars are $\pm SE$, where $SE = \sigma/\sqrt{N}$.

New incidences of skin cancer appear relatively constant in the placebo group, on average. In the beta-carotene group, the trend in mean new incidences of cancer seems to drop intially, followed by a rise from years two to four, then another drop between years four and five. However, the variance in the means, especially in the beta-carotene group, makes it difficult to argue for significant linear relationship between the number of new cancers and time.

#### ii) Provide an algebraic definition for a generalized linear marginal Poisson regression model in which the only effects are for the intercept, Year (as a continuous variable) and trt*Year. Make sure you write out the full model. Provide a table of results that you obtain from fitting the model in SAS that includes the regression coefficient estimates, empirical standard errors 95% confidence interval, and p-value for testing the hypothesis that the true parameter is zero. What do you conclude from this model about the effect of treatment?

Definition of the Poisson GLM:

1. Distribution of response: $log(\mu_{ij}) = \beta_{1} + \beta_{ij2}Year_{ij} + \beta_{ij3}Treatment_{i} \times Year_{ij}$, where
    + Year = {1, 2, 3, 4, 5}
    + Treatment = 0 if placebo, 1 if beta-carotene
2. Variance of response: $Var(Y_{ij}|X_{ij}) = \phi\mu_{ij}$
3. Correlation structure: $Corr(Y_{ij}, Y_{ik}) = \alpha_{jk}$ (compound symmetric, or exchangable; autoregressive might also be appropriate, though)

```{r simple-poisson-gee, include = FALSE}
# Fit GEE Poisson model with exchangeable (compound symmetric) correlation
simple.poisson.gee <- gee(new.count ~ year + treatment:year, # Use : syntax instead of * to suppress inclusion of treatment
                          id = id,
                          data = skin.data,
                          family = poisson,
                          corstr = "exchangeable")
simple.poisson.summary <- summary(simple.poisson.gee)

# Extract coefficient estimates
model.coefficients <- data.frame(coef(object = simple.poisson.summary))

# Calculate p-values from robust Z-scores
model.coefficients$p.value <- 2 * pnorm(q = abs(model.coefficients[, "Robust.z"]), lower.tail = FALSE)
model.coefficients$p.text <- as.character(round(model.coefficients$p.value, 3))
model.coefficients$p.text[model.coefficients$p.value < 0.001] <- "< 0.001"

# Calculate 95% C.I. from robust S.E.
model.coefficients$ci.low <- model.coefficients$Estimate - 1.96 * model.coefficients$Robust.S.E.
model.coefficients$ci.high <- model.coefficients$Estimate + 1.96 * model.coefficients$Robust.S.E.

# Extract & round only those variables we need
results.table <- data.frame(estimate = round(model.coefficients$Estimate, 3),
                            ci.low = round(model.coefficients$ci.low, 3),
                            ci.high = round(model.coefficients$ci.high, 3),
                            p.text = model.coefficients$p.text)
rownames(results.table) <- rownames(model.coefficients)
```

Results of analysis with `gee`:

| Coefficient | Covariate               | Estimate (95% CI)           | Pr($\beta = 0$) |
|:-----------:|:------------------------|:---------------------------:|:---------------:|
| $\beta_{1}$ | Intercept               | `r results.table$estimate[1]` (`r results.table$ci.low[1]`, `r results.table$ci.high[1]`) | `r results.table$p.text[1]` |
| $\beta_{2}$ | Year                    | `r results.table$estimate[2]` (`r results.table$ci.low[2]`, `r results.table$ci.high[2]`) | `r results.table$p.text[2]` |
| $\beta_{3}$ | Year $\times$ Treatment | `r results.table$estimate[3]` (`r results.table$ci.low[3]`, `r results.table$ci.high[3]`) | `r results.table$p.text[3]` |
Note the 95% confidence interval is based on the empirical (or 'robust' in `gee` parlance) standard errors.

In this analysis, treatment does not appear to influence the rate of change in the mean number of new cancers. That is, the effect of time on the mean number of new cancers does not appear significantly different between subjects in the placebo group and those in the beta-carotene group.

#### iii) Properly interpret each of the parameter estimates from your analysis in (ii) on the untransformed scale.

```{r simple-coeff-interpretation}

change <- "decreases"
if (results.table$estimate[2] > 0) {
  change <- "increases"
}

b.c.diff <- "lower"
if (results.table$estimate[3] > 0) {
  b.c.diff <- "higher"
}
```

1. $e^{\beta_{1}}$ : The expected number of new cancers in subjects before randomization (year zero). This value, $e^{\beta_{1}} =$ e^`r results.table$estimate[1]`^ = `r round(exp(results.table$estimate[1]), 3)`.
2. $e^{\beta_{2}}$ : Factor by which the average number of new cancers changes each year in the study for subjects in the placebo group. i.e. each year the average number of new cancers `r change` by a factor of $e^{\beta_{2}}$ = `r round(exp(results.table$estimate[2]), 3)`, or `r 100 * round(abs(1 - exp(results.table$estimate[2])), 3)`%, in subjects in the placebo group.
3. $e^{\beta_{3}}$ : The difference in the expected annual change in the number of new cancers between subjects in the placebo group and subjects in the beta-carotene group. That is, the annual change in average number of new cancers in subjects in the beta-carotene group is `r b.c.diff` than subjects in the placebo group by a factor of `r round(abs(1 - exp(results.table$estimate[3])), 3)`. Interpretation of this coefficient is most informative with considered in context of $\beta_{2}$. The average number of new cancers in placebo subjects `r change` by a factor of $e^{\beta_{2}}$ = `r round(exp(results.table$estimate[2]), 3)` each year. In contrast, the average number of new cancers in beta-carotene subjects is `r b.c.diff` by a factor of $e^{\beta_{2} + \beta_{3}}$ = `r round(exp(results.table$estimate[2] + results.table$estimate[3]), 3)`, or `r round(abs(1 - exp(results.table$estimate[2] + results.table$estimate[3])), 3)`%, each year.

#### iv) In the form of the results section of a scientific abstract (3 or 4 sentences), state with justification what you conclude about the effect of beta-carotene versus placebo based on your analysis.

## TODO

### 2. A secondary aim of the study was to evaluate risk factors for increased incidence of new skin cancers. To accomplish this goal,
#### i) Fit an appropriate generalized linear marginal model in SAS which includes Year (a linear effect), Treatment, Center, Age, Skin, Gender and Exposure as covariates. Make sure you fully write out your chosen model.

Definition of the Poisson GLM:

1. Distribution of response: $log(\mu_{ij}) = \beta_{1} + \beta_{ij2}Year_{ij} + \beta_{ij3}Treatment_{i} + \beta_{ij4}Center_{2i} + \beta_{ij5}Center_{3i} + \beta_{ij6}Center_{4i} + \beta_{ij7}Age_{i} + \beta_{ij8}Skin_{i} + \beta_{ij9}Gender_{i} + \beta_{ij10}Exposure_{i}$, where
$$
\begin{array}{l}
  Year_{ij} = \{1, 2, 3, 4, 5\} \\
  Treatment_{i} = 
    \begin{cases}
      0, & \text{Placebo} \\
      1, & \beta\text{-carotene}
    \end{cases} \\
  Center_{2i} = 
    \begin{cases}
      0, & \text{Not enrolled at center 2} \\
      1, & \text{Enrolled at center 2}
    \end{cases} \\
  Center_{3i} = 
    \begin{cases}
      0, & \text{Not enrolled at center 3} \\
      1, & \text{Enrolled at center 3}
    \end{cases} \\
  Center_{4i} = 
    \begin{cases}
      0, & \text{Not enrolled at center 4} \\
      1, & \text{Enrolled at center 4}
    \end{cases} \\
  Age_{i} = \text{Age at randomization} \\
  Skin_{i} = 
    \begin{cases}
      0, & \text{No burns} \\
      1, & \text{Burns}
    \end{cases} \\
  Gender_{i} = 
    \begin{cases}
      0, & \text{Female} \\
      1, & \text{Male}
    \end{cases} \\
  Exposure_{i} = \text{Number of previous skin cancers prior to randomization} \\
\end{array}
$$
2. Variance of response: $Var(Y_{ij}|X_{ij}) = \phi\mu_{ij}$
3. Correlation structure: $Corr(Y_{ij}, Y_{ik}) = \alpha_{jk}$ (compound symmetric, or exchangable; autoregressive might also be appropriate, though)

#### ii) Provide a table from your SAS output that includes parameter estimates, empirical standard errors and p-values.

```{r risk-factors-poisson-gee, include = FALSE}
# Fit GEE Poisson model with exchangeable (compound symmetric) correlation
risk.poisson.gee <- gee(new.count ~ year + treatment + center + age + skin + gender + prior.count,
                        id = id,
                        data = skin.data,
                        family = poisson,
                        corstr = "exchangeable")
risk.poisson.summary <- summary(risk.poisson.gee)

# Extract coefficient estimates & standard errors
risk.model.coefficients <- data.frame(coef(object = risk.poisson.summary))

# Calculate p-values from robust Z-scores
risk.model.coefficients$p.value <- 2 * pnorm(q = abs(risk.model.coefficients[, "Robust.z"]), lower.tail = FALSE)
risk.model.coefficients$p.text <- as.character(round(risk.model.coefficients$p.value, 3))
risk.model.coefficients$p.text[risk.model.coefficients$p.value < 0.001] <- "< 0.001"

# Extract & round only those variables we need
risk.results.table <- data.frame(estimate = round(risk.model.coefficients$Estimate, 3),
                            se = round(risk.model.coefficients$Robust.S.E., 3),
                            p.text = risk.model.coefficients$p.text)
rownames(risk.results.table) <- rownames(risk.model.coefficients)
```

| Coefficient  | Covariate     | Estimate                            | Empirical S.E.                | Pr($\beta = 0$)                   |
|:------------:|:--------------|:-----------------------------------:|:-----------------------------:|:---------------------------------:|
| $\beta_{1}$  | Intercept     | `r risk.results.table$estimate[1]`  | `r risk.results.table$se[1]`  | `r risk.results.table$p.text[1]`  |
| $\beta_{2}$  | Year          | `r risk.results.table$estimate[2]`  | `r risk.results.table$se[2]`  | `r risk.results.table$p.text[2]`  |
| $\beta_{3}$  | Treatment ($\beta$-carotene) | `r risk.results.table$estimate[3]`  | `r risk.results.table$se[3]`  | `r risk.results.table$p.text[3]`  |
| $\beta_{4}$  | Center (= 2)  | `r risk.results.table$estimate[4]`  | `r risk.results.table$se[4]`  | `r risk.results.table$p.text[4]`  |
| $\beta_{5}$  | Center (= 3)  | `r risk.results.table$estimate[5]`  | `r risk.results.table$se[5]`  | `r risk.results.table$p.text[5]`  |
| $\beta_{6}$  | Center (= 4)  | `r risk.results.table$estimate[6]`  | `r risk.results.table$se[6]`  | `r risk.results.table$p.text[6]`  |
| $\beta_{7}$  | Age           | `r risk.results.table$estimate[7]`  | `r risk.results.table$se[7]`  | `r risk.results.table$p.text[7]`  |
| $\beta_{8}$  | Skin (burns)  | `r risk.results.table$estimate[8]`  | `r risk.results.table$se[8]`  | `r risk.results.table$p.text[8]`  |
| $\beta_{9}$  | Gender (male) | `r risk.results.table$estimate[9]`  | `r risk.results.table$se[9]`  | `r risk.results.table$p.text[9]`  |
| $\beta_{10}$ | Exposure      | `r risk.results.table$estimate[10]` | `r risk.results.table$se[10]` | `r risk.results.table$p.text[10]` |

Note p-values are based on the Empirical ("Robust") S.E. Z-scores.

#### iii) In a short paragraph (<150 words), summarize the key findings from this model. 
In an analysis of several potential risk factors in the rate of new melanoma skin cancers, three factors had no significant effects while six significantly increased the number of new cancers. Specifically, when controlling for other factors, the average number of new cancers was not affected by time, treatment with $\beta$-carotene, or whether the subject had a skin type susceptible to burns. In contrast, controlling for other factors, subjects enrolled at Center #2 had, on average, higher number of new cancers than subjects enrolled at Center #1 by a factor of `r round(exp(risk.results.table$estimate[4]), 3)`. Similar differences in the average number of new cancers, controlling for other covariates, were significantly different between subjects enrolled at Center #3 and Center #1 (higher in Center #3 subjects by `r round(100 * abs(1 - exp(risk.results.table$estimate[5])), 2)`%) and between subjects enrolled at Center #4 and Center #1 (higher in Center #4 subjects by `r round(100 * abs(1 - exp(risk.results.table$estimate[6])), 2)`%). Additionally, controlling for other covariates, age at randomization also significantly affected the average number of new cancers; for each year of age, the average number of new cancers increased by a factor of `r round(exp(risk.results.table$estimate[7]), 3)`. The average number of new cancers was also higher in males than females by `r round(100 * abs(1 - exp(risk.results.table$estimate[9])), 2)`%, when controlling for other covariates. Finally, the number of cancers prior to randomization had a significant effect on the average number of new cancers when controlling for other risk factors; the average number of new cancers increased by a factor of `r round(exp(risk.results.table$estimate[10]), 3)` for each cancer measured prior to randomization. (Yes, quite a bit longer than 150 words!)

#### iv) Fit the model from 2(i) assuming there is no overdispersion (that is, the overdispersion parameter is equal to 1). Briefly describe the effect of this assumption on the results of the analysis.

This model differs from that defined in 2(i) only in the specification of variance of the mean response:

2. Variance of response: $Var(Y_{ij}|X_{ij}) = \phi\mu_{ij}$, where $\phi = 1$

The model in 2(i) does not constrain $\phi$, and, in fact, the estimated scale parameter was `r round(risk.poisson.gee$scale, 3)`.

```{r ignore-overdispersion-poisson-gee, include = FALSE}
# Fit GEE Poisson model with exchangeable (compound symmetric) correlation and assume no overdispersion
no.overdispersion.gee <- gee(new.count ~ year + treatment + center + age + skin + gender + prior.count,
                        id = id,
                        data = skin.data,
                        family = poisson,
                        corstr = "exchangeable",
                        scale.fix = TRUE,
                        scale.value = 1)
no.overdispersion.summary <- summary(no.overdispersion.gee)

# Extract coefficient estimates & standard errors
no.overdispersion.model.coefficients <- data.frame(coef(object = no.overdispersion.summary))

# Calculate p-values from robust Z-scores
no.overdispersion.model.coefficients$p.value <- 2 * pnorm(q = abs(no.overdispersion.model.coefficients[, "Robust.z"]), lower.tail = FALSE)
no.overdispersion.model.coefficients$p.text <- as.character(round(no.overdispersion.model.coefficients$p.value, 3))
no.overdispersion.model.coefficients$p.text[no.overdispersion.model.coefficients$p.value < 0.001] <- "< 0.001"

# Extract & round only those variables we need
no.overdispersion.results.table <- data.frame(estimate = round(no.overdispersion.model.coefficients$Estimate, 3),
                            se = round(no.overdispersion.model.coefficients$Robust.S.E., 3),
                            p.text = no.overdispersion.model.coefficients$p.text)
rownames(no.overdispersion.results.table) <- rownames(no.overdispersion.model.coefficients)
```

Using the following specification setting `scale.fix = TRUE` and `scale.value = 1` in the call to `gee` (`prior.count` is what I called the 'Exposure' covariate and `new.count` is the response, Y):
```{r gee-example-code, eval = FALSE, echo = TRUE}
gee.model <- gee(new.count ~ year + treatment + center + age + skin + gender + prior.count,
                 id = id,
                 data = skin.data,
                 family = poisson,
                 corstr = "exchangeable",
                 scale.fix = TRUE,
                 scale.value = 1)
```

```{r compare-naive-errors}
risk.naive <- risk.model.coefficients[, c("Naive.S.E.", "Naive.z")]
risk.naive$p.value <- 2 * pnorm(q = abs(risk.naive$Naive.z), lower.tail = FALSE)
risk.naive$p.text <- as.character(round(risk.naive$p.value, 3))
risk.naive$p.text[risk.naive$p.value < 0.001] <- "< 0.001"

no.overdispersion.naive <- no.overdispersion.model.coefficients[, c("Naive.S.E.", "Naive.z")]
no.overdispersion.naive$p.value <- 2 * pnorm(q = abs(no.overdispersion.naive$Naive.z), lower.tail = FALSE)
no.overdispersion.naive$p.text <- as.character(round(no.overdispersion.naive$p.value, 3))
no.overdispersion.naive$p.text[no.overdispersion.naive$p.value < 0.001] <- "< 0.001"
```

I found no difference in _empirical_ ("robust") parameter estimates _or_ standard errors (and therefore Z and p scores) between the original model, accounting for overdispersion (estimated $\phi = `r round(risk.poisson.gee$scale, 3)`$), and the model where I ignored overdispersion (fixed $\phi = `r no.overdispersion.gee$scale`$). However, the _model-based_ ("naive") estimates did differ slightly between the two approaches. When overdispersion was ignored ($\phi = `r no.overdispersion.gee$scale`$), model-based standard errors for coefficient estimates were all lower than the corresponding standard errors in the model accounting for overdispersion ($\phi = `r round(risk.poisson.gee$scale, 3)`$). This, in turn, resulted in greater magnitude Z-scores in model-based parameter estimates in the model ignoring overdispersion relative to the model accommodating overdispersion. 

Considering the model-based (not empirical) $\beta$ estimates' corresponding p-values:

| Coefficient  | Covariate     | $\phi = `r round(risk.poisson.gee$scale, 3)`$ | $\phi = `r round(no.overdispersion.gee$scale, 3)`$ |
|:------------:|:--------------|:---------------------------------------------:|:-----------------------------:|:------------------:|
| $\beta_{1}$  | Intercept     | `r risk.naive$p.text[1]`                      | `r no.overdispersion.naive$p.text[1]`              |
| $\beta_{2}$  | Year          | `r risk.naive$p.text[2]`                      | `r no.overdispersion.naive$p.text[2]`              |
| $\beta_{3}$  | Treatment ($\beta$-carotene) | `r risk.naive$p.text[3]`       | `r no.overdispersion.naive$p.text[3]`              |
| $\beta_{4}$  | Center (= 2)  | `r risk.naive$p.text[4]`                      | `r no.overdispersion.naive$p.text[4]`              |
| $\beta_{5}$  | Center (= 3)  | `r risk.naive$p.text[5]`                      | `r no.overdispersion.naive$p.text[5]`              |
| $\beta_{6}$  | Center (= 4)  | `r risk.naive$p.text[6]`                      | `r no.overdispersion.naive$p.text[6]`              |
| $\beta_{7}$  | Age           | `r risk.naive$p.text[7]`                      | `r no.overdispersion.naive$p.text[7]`              |
| $\beta_{8}$  | Skin (burns)  | `r risk.naive$p.text[8]`                      | `r no.overdispersion.naive$p.text[8]`              |
| $\beta_{9}$  | Gender (male) | `r risk.naive$p.text[9]`                      | `r no.overdispersion.naive$p.text[9]`              |
| $\beta_{10}$ | Exposure      | `r risk.naive$p.text[10]`                     | `r no.overdispersion.naive$p.text[10]`             |

Note that in most cases, there is no qualitative difference between models that do and do not account for overdispersion. However, there was one covariate (Skin) that had a marginally significant effect on the average number of new cancers (p = `r risk.naive$p.text[8]`) in the model where overdispersion was accounted for. In contrast, this covariate had a _significant_ effect on the average number of new cancers (p = `r no.overdispersion.naive$p.text[8]`) in the model ignoring overdispersion.

***
The R code used in this assignment can be found in a corresponding R Markdown document at [https://github.com/jcoliver/bios-684/blob/master/homework-8-oliver.Rmd](https://github.com/jcoliver/bios-684/blob/master/homework-8-oliver.Rmd).
