---
title: "Homework 4"
subtitle: "BIOS 684"
author: "Jeff Oliver"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output: 
  pdf_document:
  latex_engine: xelatex
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(echo = FALSE, dev = "pdf")
library("ggplot2")
library("nlme")
library("car") # For Anova function to get chi-squared
library("dplyr")
```

<!--
Joint Modeling of Mean and Covariance in the a Study on the Effect of Ozone on Pulmonary Function

In a study designed to examine the acute responses of normal subjects to ozone exposure, researcher randomized 20 subjects to exposure to either room air or 0.12 ppb ozone. The exposure lasted six hours. A baseline and 6 other measures of FEV1 (a measure of pulmonary function) made at hourly intervals were recorded for all study participants in the study. Subjects cycled on an exercise bike before a measure of FEV1 was obtained. The investigators were interested in determining whether changes in pulmonary function during the 6 hours of exposure were different in the ozone and room air exposed groups.

In the analyses of the data from this study, the response variable of interest was FEV1 (ml)/100. From here on out when we refer to FEV1 we mean this rescaled response, i.e., FEV1 (ml)/100. The measurement times were coded 0-6, with time=0 for the baseline measurement, time=1 for the measurement at hour 1,...,time=6 for the measurement at hour 6. For this homework, we will consider the data from hours 0, 2, 4, 6. The two exposure groups were coded 0 and 1, with group=1 denoting exposure to ozone and group=0 denoting exposure to room air. The data are in the file “ozone0246.dat” on the course web page. Each row of the data set contains the following four variables: subject ID, hour, group and the y = FEV1/100 measurements, respectively (note it is already in long format).
-->

```{r read-data}
# Read in data
ozone.long <- read.csv(file = "data/ozone0246.csv", header = FALSE)

# Set column names
colnames(ozone.long) <- c("subject", "hour", "group", "FEV1")

# Create column with hour as factor
ozone.long$cat.hour <- as.factor(ozone.long$hour)
```

### 1. Descriptive Analyses:
#### a. Describe key aspects of the longitudinal design and completeness of data.

The design is balanced and the data are complete (all ten individuals assigned to one of two groups have observations for all four of the common sampling points).

#### b. Plot the FEV1 response against hour for each subject exposed to room air (all on the first plot) and for each subject exposed to ozone (all on a second plot). Comment on any patterns in the data or other notable aspects of the data.

```{r plot-over-time, fig.height = 4}
# Vector for labeling facets
group_labels <- c("0" = "Room air", "1" = "0.12 ppb ozone")

# Plot responses over time, faceting by group
time.plot <- ggplot(data = ozone.long, mapping = aes(x = hour, y = FEV1, group = subject)) +
  geom_line() +
  ylab(label = "FEV1 (mL)/100") +
  xlab(label = "Hour") +
  facet_grid(~ group, switch = "x", labeller = labeller(group = group_labels))
time.plot
```

Forced expiratory volume from the first second of forced breath (FEV1) appeared to remain relatively constant within subjects in the room air (control) group over the six hour period of the experiment. In contrast, some individuals in the elevated ozone (treatment) group had decreasing FEV1 over the course of the experiment. Given this variation in response within the ozone treatment, it looks like the between-subject variance in FEV1 in the ozone treatment likely increases through time.

#### c. Obtain the mean FEV1 value at each hour of measurement for ozone and room air subjects separately. Plot the means against hour. Comment on the pattern of change in mean FEV1 with hour for ozone and room air subjects.

```{r means-plot, fig.height = 4}
# Extract mean, sd, and counts
desc.stats <- ozone.long %>% 
  group_by(group, hour) %>% 
  summarize(mean = mean(FEV1), sd = sd(FEV1), count = n())

# Convert to a df and calculate SE
desc.stats <- data.frame(desc.stats)
desc.stats$group <- as.factor(desc.stats$group)
desc.stats$se <- desc.stats$sd / sqrt(desc.stats$count)


# Plot both control and treatment group means
means.plot <- ggplot(data = desc.stats, mapping = aes(x = hour, y = mean, color = group, group = group)) +
  geom_point() +
  geom_line() +
  geom_errorbar(data = desc.stats, mapping = aes(x = hour, ymin = (mean - se), ymax = (mean + se)), width = 0.2) +
  scale_color_manual(name = "Treatment", values = c("cadetblue", "darkred"), labels = group_labels) +
  ylab(label = "FEV1 (mL)/100") +
  xlab(label = "Hour")
means.plot
```
Note error bars show $\pm 1SE$ where $SE = \sigma/\sqrt{n}$.

The mean FEV1 in control subjects appears fairly constant over time (there is a suggestion of a dip between the baseline and 2 hour measurement, followed by an increase between 2 and 4 and 4 and 6 hour timepoints, but unlikely to be significant), while the mean FEV1 appears to decrease in the treatment group. Also of note is that between-subjects variance is the control group appears constant, while between-subject variance in the treatment group appears to increase with time.

### 2. Fitting a “Maximal” Model and Evaluating Variance-Covariance Structure:
#### a. Define a reasonable “maximal” mean model for this study. Fit this model using an unstructured variance-covariance matrix. Comment on the variance structure and on the correlation structure. What simplified variance-covariance structure(s) might be reasonable? Justify your answer.

A reasonable "maximal" mean model for this study would treat time as categorical:
$$
Y_{ij} = \beta_{1}X_{1ij} + \beta_{2}X_{2ij} + \beta_{3}X_{3ij} + \beta_{4}X_{4ij} + \beta_{5}X_{5ij} + \\
\beta_{7}(X_{2ij} \times X_{3ij}) + \beta_{8}(X_{2ij} \times X_{4ij}) + \beta_{9}(X_{2ij} \times X_{5ij}) + \varepsilon_{ij} 
$$
Where:

$$
\begin{array}{l}
  X_{1ij} = 1 \\
    X_{2ij} = 
    \begin{cases}
      0, & \text{for room air control} \\
      1, & \text{for ozone treatment}
    \end{cases} \\
  X_{3ij} = 1 \text{ if measurement is for hour 2, 0 otherwise} \\
  X_{4ij} = 1 \text{ if measurement is for hour 4, 0 otherwise} \\
  X_{5ij} = 1 \text{ if measurement is for hour 6, 0 otherwise}
\end{array}
$$

```{r unstructured-model}
# Build gls unstructured variance model
unstructured.model <- gls(FEV1 ~ group + cat.hour + group * cat.hour,
                     data = ozone.long,
                     correlation = corSymm(form = ~1|subject),
                     weights = varIdent(form = ~1|cat.hour),
                     method = "REML")

unstructured.summary <- summary(unstructured.model)

# Build storage object for subsequent model comparison
var.models <- data.frame(model.name = c("Unstructured", "Compound Symmetric", "Heterogeneous Compound Symmetric", 
                                         "First-order Autoregressive", "Heterogeneous first-order Autoregressive"),
                         model.code = c("un", "cs", "hcs", "ar", "har"),
                         num.params = NA,
                         lnl = NA,
                         aic = NA)
rownames(var.models) <- var.models$model.code
var.models["un", "num.params"] <- (nlevels(ozone.long$cat.hour)*(nlevels(ozone.long$cat.hour) + 1)) / 2
var.models["un", "lnl"] <- unstructured.summary$logLik[1]
var.models["un", "aic"] <- unstructured.summary$AIC[1]
```

The unstructured variance-covariance matrix has $n(n+1)/2$ parameters ($n$ variances and $n(n-1)/2$ pairwise covariances). In this experiment, this means there are `r nlevels(ozone.long$cat.hour)*(nlevels(ozone.long$cat.hour) + 1)/2` parameters estimated for the variance-covariance matrix. A reasonable simplified variance-covariance structure would be a first-order autoregressive structure. This is justified because the measurements were taken at the same times and there are only a few measurement points, thus the maximal decline in covariance will be an order of 3:

$$
Cov(Y_{i}) = \sigma^2
\begin{pmatrix}
1 & \rho & \rho^2 & \rho^3 \\
\rho & 1 & \rho & \rho^2 \\
\rho^2 & \rho & 1 & \rho \\
\rho^3 & \rho^2 & \rho & 1 \\
\end{pmatrix}
$$

This has the benefit of only 2 estimated parameters ($\sigma$ and $\rho$).

However, the variance, at least in the treatment group, does _not_ appear to be constant over time (it seems to increase), so another appropriate variance-covariance model, heterogeneous first-order auto-regressive, would relax the assumption of constant variance:

$$
Cov(Y_{i}) = 
\begin{pmatrix}
\sigma_{1}^2 & \rho & \rho^2 & \rho^3 \\
\rho & \sigma_{2}^2 & \rho & \rho^2 \\
\rho^2 & \rho & \sigma_{3}^2 & \rho \\
\rho^3 & \rho^2 & \rho & \sigma_{4}^2 \\
\end{pmatrix}
$$
This model has `r nlevels(ozone.long$cat.hour) + 1` ($n + 1$) parameters.

#### b. Keeping the same maximal mean model, evaluate whether your suggestion(s) for the variance-covariance structure from question 2a as well as the following models for the variance-covariance structure provide an adequate fit to the data compared with an unstructured variance-covariance:

1. compound symmetry
2. heterogeneous compound symmetry (hint use: `type=csh` in `proc mixed`) 
3. 1^st^-order autoregressive
4. heterogeneous 1^st^-order autoregressive (hint use: `type=arh(1)` in `proc mixed`)

```{r structured-covariance-models}
# Compound symmetry
cs.model <- gls(FEV1 ~ group + cat.hour + group * cat.hour,
                     data = ozone.long,
                     correlation = corCompSymm(form = ~1|subject),
                     weights = varIdent(form = ~1),
                     method = "REML")

cs.summary <- summary(cs.model)
var.models["cs", "num.params"] <- 2
var.models["cs", "lnl"] <- cs.summary$logLik[1]
var.models["cs", "aic"] <- cs.summary$AIC[1]

# Heterogeneous compound symmetry
hcs.model <- gls(FEV1 ~ group + cat.hour + group * cat.hour,
                     data = ozone.long,
                     correlation = corCompSymm(form = ~1|subject),
                     weights = varIdent(form = ~1|cat.hour),
                     method = "REML")

hcs.summary <- summary(hcs.model)
var.models["hcs", "num.params"] <- nlevels(ozone.long$cat.hour) + 1
var.models["hcs", "lnl"] <- hcs.summary$logLik[1]
var.models["hcs", "aic"] <- hcs.summary$AIC[1]

# First-order autoregressive
ar.model <- gls(FEV1 ~ group + cat.hour + group * cat.hour,
                     data = ozone.long,
                     correlation = corAR1(form = ~1|subject),
                     weights = varIdent(form = ~1),
                     method = "REML")

ar.summary <- summary(ar.model)
var.models["ar", "num.params"] <- 2
var.models["ar", "lnl"] <- ar.summary$logLik[1]
var.models["ar", "aic"] <- ar.summary$AIC[1]

# Heterogeneous first-order autoregressive
har.model <- gls(FEV1 ~ group + cat.hour + group * cat.hour,
                     data = ozone.long,
                     correlation = corAR1(form = ~1|subject),
                     weights = varIdent(form = ~1|cat.hour),
                     method = "REML")

har.summary <- summary(har.model)
var.models["har", "num.params"] <- nlevels(ozone.long$cat.hour) + 1
var.models["har", "lnl"] <- har.summary$logLik[1]
var.models["har", "aic"] <- har.summary$AIC[1]
```

#### Using likelihood ratio tests and the AIC criterion as appropriate, identify a model for the variance-covariance structure that provides a good fit to the data. Provide estimates for the parameters used in defining this variance-covariance model. Also provide estimated variance-covariance and correlation matrices.

I start by comparing each model to the unstructured variance-covariance model (_ln(L)_ = `r round(var.models["un", "lnl"], 3)`). In all cases, log-likelihoods are from a REML model. P-values are based on a comparison of $2\Delta ln(L)$ to a $\chi^2$ distribution with degrees of freedom equal to the difference in the number of covariance parameters between the model in question and the unstructured covariance model.

| Model | # Covariance Parameters | $ln(L)$ | $2\Delta ln(L)$ | $p$ |
|:------|:-----------------------:|:-------:|:---------------:|:---:|
| `r var.models["cs", "model.name"]` | `r var.models["cs", "num.params"]` | `r round(var.models["cs", "lnl"], 3)` | `r round(2 * (var.models["un", "lnl"] - var.models["cs", "lnl"]), 3)` | $`r signif(pchisq(q = 2 * (var.models["un", "lnl"] - var.models["cs", "lnl"]), df = (10 - var.models["cs", "num.params"]), lower.tail = FALSE), 3)`$ |
| `r var.models["hcs", "model.name"]` | `r var.models["hcs", "num.params"]` | `r round(var.models["hcs", "lnl"], 3)` | `r round(2 * (var.models["un", "lnl"] - var.models["hcs", "lnl"]), 3)` | $`r round(pchisq(q = 2 * (var.models["un", "lnl"] - var.models["hcs", "lnl"]), df = (10 - var.models["hcs", "num.params"]), lower.tail = FALSE), 3)`$ |
| `r var.models["ar", "model.name"]` | `r var.models["ar", "num.params"]` | `r round(var.models["ar", "lnl"], 3)` | `r round(2 * (var.models["un", "lnl"] - var.models["ar", "lnl"]), 3)` | $`r round(pchisq(q = 2 * (var.models["un", "lnl"] - var.models["ar", "lnl"]), df = (10 - var.models["ar", "num.params"]), lower.tail = FALSE), 3)`$ |
| `r var.models["har", "model.name"]` | `r var.models["har", "num.params"]` | `r round(var.models["har", "lnl"], 3)` | `r round(2 * (var.models["un", "lnl"] - var.models["har", "lnl"]), 3)` | $`r round(pchisq(q = 2 * (var.models["un", "lnl"] - var.models["har", "lnl"]), df = (10 - var.models["har", "num.params"]), lower.tail = FALSE), 3)`$ |

The two simplest models, the compound symmetric and the first-order autoregressive, both provided a significantly worse fit than the more complex unstructured model (both had $p \leq 0.001$ in likelihood ratio test). In contrast, the two models that did not assume constant variance, the heterogeneous compound symmetric and the heterogeneous first-order autoregressive, did not perform significantly worse than the unstructured model. These simpler models offer two potential alternatives to the parameter-rich unstructured covariance model. To compare these models, we need to use AIC rather than a likelihood ratio test because they are not nested models (one is not a special case of the other). 

| Model | AIC |
|:------|:---:|
| `r var.models["hcs", "model.name"]` | `r round(var.models["hcs", "aic"], 3)` |
| `r var.models["har", "model.name"]` | `r round(var.models["har", "aic"], 3)` |

Because the heterogeneous first-order autoregressive model has a lower AIC score ($\Delta AIC = `r round(var.models["hcs", "aic"] - var.models["har", "aic"], 3)`$), that is the model that provides a sufficient fit to the model without over-parameterizing the variance-covariance structure.

The $\beta$ coefficients estimated under the heterogeneous first-order autoregressive variance-covariance model are:

| Coefficient | Estimate | Description |
|:-----------:|:--------:|:------------|
| $\beta_{1}$ | `r round(har.summary$coefficients[[1]], 3)` | Intercept | 
| $\beta_{2}$ | `r round(har.summary$coefficients[[2]], 3)` | Group | 
| $\beta_{3}$ | `r round(har.summary$coefficients[[3]], 3)` | Hour 2 | 
| $\beta_{4}$ | `r round(har.summary$coefficients[[4]], 3)` | Hour 4 | 
| $\beta_{5}$ | `r round(har.summary$coefficients[[5]], 3)` | Hour 6 | 
| $\beta_{6}$ | `r round(har.summary$coefficients[[6]], 3)` | Group $\times$ hour 2 | 
| $\beta_{7}$ | `r round(har.summary$coefficients[[7]], 3)` | Group $\times$ hour 4 | 
| $\beta_{8}$ | `r round(har.summary$coefficients[[8]], 3)` | Group $\times$ hour 6 | 

And the variance-covariance matrix for this model is:

```{r har-var-covar}
har.covar <- getVarCov(obj = har.model)
har.covar <- round(har.covar, 3)
```

$$
Cov(Y_{i}) = 
\begin{pmatrix}
`r har.covar[1, 1]` & `r har.covar[1, 2]` & `r har.covar[1, 3]` & `r har.covar[1, 4]` \\
`r har.covar[2, 1]` & `r har.covar[2, 2]` & `r har.covar[2, 3]` & `r har.covar[2, 4]` \\
`r har.covar[3, 1]` & `r har.covar[3, 2]` & `r har.covar[3, 3]` & `r har.covar[3, 4]` \\
`r har.covar[4, 1]` & `r har.covar[4, 2]` & `r har.covar[4, 3]` & `r har.covar[4, 4]` \\
\end{pmatrix}
$$

### 3. Analysis of Response Profiles: Fit the usual model for the analysis of mean profiles using room air exposure as the reference level for group and baseline as the reference group for time. Use the variance-covariance structure identified in your answer to question 2b. Based on this model:

```{r}
har.ml.model <- gls(FEV1 ~ group + cat.hour + group * cat.hour,
                    data = ozone.long,
                    correlation = corAR1(form = ~1|subject),
                    weights = varIdent(form = ~1|cat.hour),
                    method = "ML")

har.ml.summary <- summary(har.ml.model)
har.ml.anova <- Anova(har.ml.model)
coincide.chisq <- round(sum(har.ml.anova$Chisq[c(1, 3)]), 3)
coincide.df <- sum(har.ml.anova$Df[c(1, 3)])
coincide.p <- round(pchisq(q = sum(har.ml.anova$Chisq[c(1, 3)]), df = sum(har.ml.anova$Df[c(1, 3)]), lower.tail = FALSE), 3)
coincide.text <- "do"
if (coincide.p >= 0.05) {
  coincide.text <- "do not"
}
parallel.chisq <- round(har.ml.anova$Chisq[3], 3)
parallel.df <- har.ml.anova$Df[3]
parallel.p <- har.ml.anova$`Pr(>Chisq)`[3]
parallel.text <- "do"
if (parallel.p >= 0.05) {
  parallel.text <- "do not"
}
```

#### a. Test the null hypothesis that the pattern of means over hours is identical (coincides) for the two exposure groups. What do you conclude?

In a test to see if the response pattern coincides for the two exposure groups, we perform a joint test of the effect of group and group $\times$ time interaction. In the model, using maximum likelihood estimation (not REML), I tested $H_{0}: \beta_{2} = \beta_{6} = \beta_{7} = \beta_{8} = 0$. The test statistic _W^2^_ = `r coincide.chisq`. With `r coincide.df` degrees of freedom, this indicates we _`r coincide.text`_ reject the hypothesis that the responses of the two groups coincide (p = `r coincide.p`).

#### b. Test the null hypothesis that the mean response profiles of the two groups are parallel. What do you conclude?

In this case we are testing only the coefficients for the group $\times$ time interactions, or $H_{0}: \beta_{6} = \beta_{7} = \beta_{8} = 0$. For this test, _W^2^_ = `r parallel.chisq` and there are `r parallel.df` degrees of freedom. In this case, we _`r parallel.text`_ reject the null hypothesis of parallel mean response profiles of the two groups (p = `r round(parallel.p, 3)`).

### 4. Fitting a Linear Model in Time: Fit a model that includes hour as a continuous variable, group and their interaction. Use the model for the variance-covariance structure that you identified in question 2b.

```{r continuous-time}
linear.model <- gls(FEV1 ~ group + hour + group * hour,
                    data = ozone.long,
                    correlation = corAR1(form = ~1|subject),
                    weights = varIdent(form = ~1|hour),
                    method = "REML")
linear.summary <- summary(linear.model)
coeffs.95CI <- round(confint(object = linear.model), 3)
```

Using the heterogeneous first-order autoregressive variance-covariance structure, the response for the mean is:

$$
Y_{ij} = \beta_{1} + \beta_{2}Group_{i} + \beta_{3}t_{ij} + \beta_{4}(Group_{i} \times t_{ij}) + \varepsilon_{ij}
$$

#### a. What is the estimated rate of change in mean response for the room air group?

The estimated rate of change in the mean response for subjects in the room air group is $\beta_{3}$ = `r round(linear.summary$coefficients[[3]], 3)`. That is, for every hour in the experiment, we expect the FEV1 to decrease by $`r abs(round(linear.summary$coefficients[[3]], 3))/100`$ mL in control group subjects.

#### b. What is the estimated rate of change in mean response for the ozone group?

The estimated rate of change in mean response for subjects in the ozone group is $\beta_{3} + \beta_{4}$ = `r round((linear.summary$coefficients[[3]] + linear.summary$coefficients[[4]]), 3)`. That is, for every hour in the experiment, we expect the FEV1 to decrease by `r abs(round((linear.summary$coefficients[[3]] + linear.summary$coefficients[[4]]), 3))/100` mL in subjects exposed to ozone.

#### c. Test the hypothesis that the rates of change in mean response are identical in the two groups. What do you conclude?

The rates of change in the mean response are not identical in the two groups. We reject this hypothesis ($H_{0}: \beta_{4} = 0$) at p = `r round(linear.summary$tTable[4, 4], 3)`.

#### d. What is the estimated difference in rate of mean change between the two groups? By calculating a 95% confidence interval for this difference, identify what are plausible values for the underlying true difference.

The estimated difference in the rate of mean change between the two groups is $\beta_{4}$ = `r round(linear.summary$coefficients[[4]], 3)`. The 95% confidence interval for this estimate is [`r coeffs.95CI[4, 1]`, `r coeffs.95CI[4, 2]`].

### 5. Evaluating the Fit of a Linear Model in Time: Does a model with a linear trend in hour for each exposure group adequately describe the pattern of change in the two groups? Justify your answer with appropriate statistical analysis. (Hint: compare it with quadratic model and statured model)

```{r continuous-time-comparisons}
# Linear model with ML instead of REML
linear.ml.model <- gls(FEV1 ~ group + hour + group * hour,
                    data = ozone.long,
                    correlation = corAR1(form = ~1|subject),
                    weights = varIdent(form = ~1|hour),
                    method = "ML")
linear.ml.summary <- summary(linear.ml.model)

# Center time before analysis
ozone.long$hour.centered <- ozone.long$hour - mean(as.numeric(levels(ozone.long$cat.hour)))
ozone.long$hour.c.sq <- ozone.long$hour.centered^2
quad.model <- gls(FEV1 ~ group + hour.centered + hour.c.sq + group * hour.centered + group * hour.c.sq,
                    data = ozone.long,
                    correlation = corAR1(form = ~1|subject),
                    weights = varIdent(form = ~1|hour.centered),
                    method = "ML")
quad.summary <- summary(quad.model)

# Extract info for LRT between parametric models
linear.v.quad.lnl <- 2 * (quad.summary$logLik[1] - linear.ml.summary$logLik[1])
linear.v.quad.df <- length(quad.summary$coefficients) - length(linear.ml.summary$coefficients)
linear.v.quad.p <- round(pchisq(q = linear.v.quad.lnl, df = linear.v.quad.df, lower.tail = FALSE), 3)

# Extract info for AIC between linear and saturated
linear.v.sat.aic <- round(har.ml.summary$AIC - linear.summary$AIC, 3)
```

The quadratic model for the mean is: 

$$
Y_{ij} = \beta_{1} + \beta_{2}Group_{i} + \beta_{3}t_{ij} + \beta_{4}t_{ij}^2 + \beta_{5}(Group_{i} \times t_{ij}) + \beta_{6}(Group_{i} \times t_{ij}^2) + \varepsilon_{ij}
$$
Where $t_{ij}$ is centered on the mean.

In a comparison of linear and quadradic models, the quadratic model _does not_ result in a significant improvement in log-likelihood score ($2\Delta lnL = `r round(linear.v.quad.lnl, 3)`$, df = `r linear.v.quad.df`, $p = `r linear.v.quad.p`$). Therefore, in this comparison, the linear model is sufficient to describe the change in FEV1 over time when compared with a quadratic model.

Recall the saturated model, where time is a categorical variable:

$$
Y_{ij} = \beta_{1}X_{1ij} + \beta_{2}X_{2ij} + \beta_{3}X_{3ij} + \beta_{4}X_{4ij} + \beta_{5}X_{5ij} + \\
\beta_{7}(X_{2ij} \times X_{3ij}) + \beta_{8}(X_{2ij} \times X_{4ij}) + \beta_{9}(X_{2ij} \times X_{5ij}) + \varepsilon_{ij} 
$$

and:

$$
\begin{array}{l}
  X_{1ij} = 1 \\
    X_{2ij} = 
    \begin{cases}
      0, & \text{for room air control} \\
      1, & \text{for ozone treatment}
    \end{cases} \\
  X_{3ij} = 1 \text{ if measurement is for hour 2, 0 otherwise} \\
  X_{4ij} = 1 \text{ if measurement is for hour 4, 0 otherwise} \\
  X_{5ij} = 1 \text{ if measurement is for hour 6, 0 otherwise}
\end{array}
$$

To compare this saturated model to the simpler linear model, I used AIC, as the simple model is not nested within the saturated model. While the saturated model had a slightly better maximum likelihood score ($-lnL = `r -round(har.ml.summary$logLik[1], 3)`$) than the linear model ($-lnL = `r -round(linear.ml.summary$logLik[1], 3)`$), the linear model had a substantially better AIC score ($\Delta AIC = `r linear.v.sat.aic`$) than the saturated model (largely due to the difference in the number of estimated parameters between the two models). Thus, the simpler linear model remains sufficient for explaining the response in both groups over time.

### 6. Summarizing the Key Results and Conclusions: Write a brief structured abstract (maximum 200 words) summarizing the objective, methods, results and conclusions that might be drawn concerning exposure differences in patterns of pulmonary function over time.

```{r abstract}
# Get confidence intervals
linear.ci <- confint(object = linear.model)

# Accumulate statistics for abstract

# Magnitude and directionality of change over time for control group
control.change <- round(linear.summary$coefficients["hour"][1], 2)
control.change.dir <- "increase"
if (control.change < 0) {
  control.change.dir <- "decrease"
  control.change <- abs(control.change)
}

# 95% CI for the change over time for control group
control.change.ci <- round(linear.ci["hour", ], 2)

# Magnitude and directionality of change over time for treatment group rel. to control
ozone.change <- round(linear.summary$coefficients["group:hour"][1], 2)
ozone.change.dir <- "increase"
if (ozone.change < 0) {
  ozone.change.dir <- "decrease"
  ozone.change <- abs(ozone.change)
}

# 95% CI for the change over time for control group
ozone.change.ci <- round(linear.ci["group:hour", ], 2)

# And convert them all to mL
control.change <- control.change/100
control.change.ci <- control.change.ci/100
ozone.change <- ozone.change/100
ozone.change.ci <- ozone.change.ci/100
```

**Objective**  
The purpose of this study is to investigate the short-term effects of ozone exposure on pulmonary function.

**Methods**  
Twenty subjects were randomized into one of two groups: a control group exposed to room air and a treatment group exposed to air with 0.12 ppb ozone. Subjects cycled on a stationary bike before pulmonary function was quantified by FEV1. FEV1 was measured before exposure (baseline) and every hour thereafter, for six hours. For the purposes of this analysis, only measurements at 0, 2, 4, and 6 hours are considered. We investigated the effect of ozone exposure on pulmonary function with a time-continuous model.

**Results**  
In the control group, we found no significant change in FEV1 over the course of the experiment; the expected change in FEV1 per hour was a $`r control.change`$ mL (95% CI: [`r control.change.ci[[1]]`, `r control.change.ci[[2]]`]) `r control.change.dir` (note the 95% CI includes zero). In the treatment group, where subjects were exposed to elevated ozone levels, there was a significant `r ozone.change.dir` in FEV1. Those subjects in the treatment group had an `r ozone.change.dir` of $`r ozone.change`$ mL per hour (95% CI: [`r ozone.change.ci[[1]]`, `r ozone.change.ci[[2]]`]) relative to subjects in the control group.

**Conclusions**  
Exposure to ozone can lead to significant short-term decreases in pulmonary function. 


***
The R code used in this assignment can be found in a corresponding R Markdown document at [https://github.com/jcoliver/bios-684/blob/master/homework-4-oliver.Rmd](https://github.com/jcoliver/bios-684/blob/master/homework-4-oliver.Rmd).
