---
title: "Homework 1"
subtitle: "BIOS 684"
author: "Jeff Oliver"
date: "05 September 2017"
output: 
  pdf_document:
  latex_engine: xelatex
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

<!--
The dataset “hers.txt” available in the “Datasets” folder of the course webpage includes data from the Heart and Estrogen/Progestin Replacement study (HERS), a randomized clinical trial investigating the efficacy of hormone replacement therapy (HT) for secondary prevention of coronary heart disease. For each study subject, in addition to HT, investigators measured a set of known CHD risk factors, including LDL and HDL cholesterol levels, age, bmi, as well as other important subject characteristics, such as statin use. The dataset on the course webpage includes a row of data for each of 2747 subjects enrolled in the study. In each row, the variables are ln(LDL), age, BMI, and statin use (1=yes, 0=no).
-->

```{r echo = FALSE}
# Read in data
hers.data <- read.delim(file = "data/hers.txt",
                        header = FALSE,
                        sep = " ")
# Add column names
colnames(hers.data) <- c("ln.LDL", "age", "bmi", "statin")
```

### 1. Provide a plot of ln(LDL) against each of age, BMI, and statin use. Very briefly, provide a qualitative characterization of the associations that you see.

```{r echo = FALSE}
# Plot all three potential relationships as appropriate
op <- par(mfrow = c(1,1))
par(mfrow = c(2, 2),
    mar = c(3, 3, 2, 1) + 0.1,
    las = 1,
    mgp = c(1.75, 0.6, 0),
    cex.axis = 0.8)
plot(ln.LDL ~ age, 
     data = hers.data,
     main = "LDL vs Age",
     xlab = "Age (years)",
     ylab = "ln(LDL)")
box("figure")
plot(ln.LDL ~ bmi, 
     data = hers.data,
     main = "LDL vs BMI",
     xlab = "BMI",
     ylab = "ln(LDL)")
box("figure")
boxplot(ln.LDL ~ statin, 
        data = hers.data,
        main = "LDL vs. Statin Use",
        xlab = "Statin Use",
        ylab = "ln(LDL)",
        names = c("No", "Yes"))
box("figure")
par(op)
```

It is difficult to see much of a relationship between the two continuous variables (age and body mass index) and the cholestorol level. There might be higher cholesterol in those subjects that were _not_ using statin, but there seems to be considerable overlap in the distribution between the two subject groups. The upper boundary of the age distribution is a little disturbing.

### 2. Use PROC REG in SAS to fit a regression model to describe each of the three associations plotted in response to question 1. Provide a table showing, for each parameter in the mean part of the model, the estimated parameter, standard error and p-value from a test that the true parameter is equal to zero. Provide a brief quantitative characterization of the two [_sic_] associations. [HINT: The parameters in the mean part of the model in the class notes are the $\beta$’s].

```{r echo = FALSE}
# Vector to iterate over all predictors
predictors <- c("age", "bmi", "statin")

# Matrix to store results of univariate analyses
model.stats <- matrix(data = NA,
                      nrow = length(predictors),
                      ncol = 3)
rownames(model.stats) <- predictors
colnames(model.stats) <- c("param.est", "std.error", "p.value")

# Iterate over each predictor, running model and extracting values from coefficients element
for (predictor in predictors) {
  univar.model <- lm(paste0("ln.LDL ~ ", predictor), data = hers.data)
  univar.summary <- summary(univar.model)
  model.stats[predictor, ] <- univar.summary$coefficients[2, c(1, 2, 4)]
}

# Deal with formatting of p-values, some of which will end up equal to zero after rounding;
# set these cases as equal to the string "< 0.001"
univar.p.values <- model.stats[, 3]
univar.p.values <- round(univar.p.values, 3)
univar.p.values[univar.p.values == 0] <- NA
univar.p.values <- as.character(univar.p.values)
univar.p.values[is.na(univar.p.values)] <- "< 0.001"
names(univar.p.values) <- rownames(model.stats)
```

Results of univariate linear models, measuring individual associations between ln(LDL) and each response (age, body mass index, and statin use):

|Predictor|Parameter Estimate|Standard Error|P-value|
|---------|:----------------:|:------------:|:-----:|
| `r rownames(model.stats)[1]` | `r round(model.stats[1, 1], 3)` | `r round(model.stats[1, 2], 3)` | `r univar.p.values["age"]` |
| `r rownames(model.stats)[2]` | `r round(model.stats[2, 1], 3)` | `r round(model.stats[2, 2], 3)` | `r univar.p.values["bmi"]` |
| `r rownames(model.stats)[3]` | `r round(model.stats[3, 1], 3)` | `r round(model.stats[3, 2], 3)` | `r univar.p.values["statin"]` |

The associations between each response variable and cholesterol are significant at the $\alpha = 0.05$ level. Comparing the univariate models, statin use had the largest effect, with statin use reducing ln(LDL) by `r abs(round(model.stats[3, 1], 2))` units (I don't know how LDL was measured, so I wasn't sure about units). Body mass index had a positive relationship on ln(LDL), while age had a negative impact on ln(LDL).

### 3. Use PROC REG is SAS to fit a single model describing the multivariable association between ln(LDL) and age, BMI, and statin use.
#### 3.1 Write down a complete algebraic definition of the model being fitted, including assumptions.

The multivariate regression model is: 

$$
ln(LDL) = \beta_{1}age + \beta_{2}bmi + \beta_{3}statin
$$

Or, more abstractly:

$$
Y_{j} = \beta_{0} + \beta_{1}X_{j1} + \beta_{2}X_{j2} + \beta_{3}X_{j3} + \epsilon_{j}
$$

Where $Y_{j}$ is the observed value of ln(LDL) in the _j_^th^ subject and $X_{j1}$, $X_{j2}$, and $X_{j3}$, are the observed values of age, body mass index, and statin use, repsectively, in the _j_^th^ subject. The model for the mean is:

$$
E(Y) = X\beta
$$

Expanded, this is:
$$
\left( 
\begin{array}{c}
E(Y_{1}|X_{11}, X_{12}, X_{13}) \\ E(Y_{2}|X_{21}, X_{22}, X_{23}) \\ \vdots \\ E(Y_{n}|X_{n1}, X_{n2}, X_{n3})
\end{array} 
\right)
=
\begin{pmatrix}
1 & X_{11} & X_{12} & X_{13} \\
1 & X_{21} & X_{22} & X_{23} \\
\vdots & \vdots & \vdots & \vdots \\
1 & X_{n1} & X_{n2} & X_{n3}
\end{pmatrix}
\left ( 
\begin{array}{c}
\beta_{0} \\ \beta_{1} \\ \beta_{2} \\ \beta_{3}
\end{array}
\right)
=
\begin{pmatrix}
\beta_{0} + \beta_{1}X_{11} + \beta_{2}X_{12} + \beta_{3}X_{13} \\
\beta_{0} + \beta_{1}X_{21} + \beta_{2}X_{22} + \beta_{3}X_{23} \\
\vdots \\
\beta_{0} + \beta_{1}X_{n1} + \beta_{2}X_{n2} + \beta_{3}X_{n3} \\
\end{pmatrix}
$$

This model for the mean assumes:

+ **Independence** among subjects, i.e. the value of $Y_{j}$ is not related to the values of $Y_{i}$ (when $i \neq j$).
+ The relationships between response (ln(LDL)) and predictors (age, body mass index, and statin use) are **linear**.
+ The response variable ln(LDL) follows a **normal distribution**.
* The variation in ln(LDL) should be independent of the value of the predictor variables, i.e. variance should be **homoscedastic**.
+ Individuals are sampled at the population at **random**.

#### 3.2 Based on your algebraic definition, show how ln(LDL) is associated with age, BMI, and statin use.
```{r echo = FALSE}
# Run multivariate analysis, including all three predictor variables
multi.model <- lm(ln.LDL ~ age + bmi + statin, data = hers.data)
multi.summary <- summary(multi.model)
coeffs <- multi.summary$coefficients

# Deal with p-values as above (correcting for rounding down to zero)
multi.p.values <- coeffs[, 4]
multi.p.values <- round(multi.p.values, 3)
multi.p.values[multi.p.values == 0] <- NA
multi.p.values <- as.character(multi.p.values)
multi.p.values[is.na(multi.p.values)] <- "< 0.001"
names(multi.p.values) <- rownames(coeffs)
```
After running multiple linear regression with the `lm` function in R, the model is:  
  
_ln(LDL)_ = `r round(coeffs["(Intercept)", 1], 3)` + `r round(coeffs["age", 1], 3)` _age_ + `r round(coeffs["bmi", 1], 3)` _bmi_ + `r round(coeffs["statin", 1], 3)` _statin_

#### 3.3 Provide a table showing, for each parameter in the mean part of the model, the estimated parameter, standard error and p-value from a test that the true parameter is equal to zero. Provide a quantitative interpretation of the estimates and 95% confidence intervals for the parameters in this model.

```{r echo = FALSE}
# Calculate 95% confidence intervals
ci.95 <- confint(multi.model)
```

|Parameter|Parameter Estimate|Standard Error|Pr($\neq0$)|95% Confidence Interval|
|---------|:----------------:|:------------:|:---------:|:---------------------:|
| Intercept | `r round(coeffs["(Intercept)", 1], 3)` | `r round(coeffs["(Intercept)", 2], 3)` | `r multi.p.values["(Intercept)"]` | `r round(ci.95["(Intercept)", 1], 3)` - `r round(ci.95["(Intercept)", 2], 3)`|
| Age | `r round(coeffs["age", 1], 3)` | `r round(coeffs["age", 2], 3)` | `r multi.p.values["age"]` | `r round(ci.95["age", 1], 3)` - `r round(ci.95["age", 2], 3)`|
| BMI | `r round(coeffs["bmi", 1], 3)` | `r round(coeffs["bmi", 2], 3)` | `r multi.p.values["bmi"]` | `r round(ci.95["bmi", 1], 3)` - `r round(ci.95["bmi", 2], 3)`|
| Statin use | `r round(coeffs["statin", 1], 3)` | `r round(coeffs["statin", 2], 3)` | `r multi.p.values["statin"]` | `r round(ci.95["statin", 1], 3)` - `r round(ci.95["statin", 2], 3)`|


```{r echo = FALSE}
# Calculate the effect of predictors on LDL (instead of ln(LDL))
exp.coeffs <- exp(x = coeffs[, 1])
effects <- abs(1 - exp.coeffs) * 100
```

+ Age: Increasing age led to a reduction in LDL, with LDL decreasing approximately `r round(effects["age"], 2)`% per year, when body mass index and statin use are held constant. 
+ BMI: One unit increase in body mass index decreased LDL levels by `r round(effects["bmi"], 2)`%, when age and statin use are held constant.
+ Statin use: Subjects on statin had LDL levels `r round(effects["statin"], 2)`% lower than those subjects not on statin, while controlling for age and body mass index.

### 4. Based on the results of the analyses conducted in response to questions 1 to 3, summarize the methods, results and interpretation of this study and analysis in a brief structured paragraph in a form that is informative and suitable for an abstract for submission to a conference (which has a 250 word limit). [HINT: It would be good to provide information about the association of LDL with each predictor, not ln(LDL)].

**Introduction.** There are a variety therapuetic treatments that may influence the levels of LDL cholesterol, including the use of statins. Here we investigate the effects of statin use on LDL cholesterol levels, while accounting for age and body mass index.

**Methods.** We used multiple linear regression to simultaneously assess the effects of age, body mass index, and statin use on `r nrow(hers.data)` subjects from the Heart and Estrogen/Progestin Replacement study.

**Results.** All tested predictors had a significant effect on LDL cholesterol levels. Increasing age led to a reduction in LDL, with LDL decreasing approximately `r round(effects["age"], 2)`% per year, when body mass index and statin use are held constant. In contrast, body mass index had a negative effect on LDL; one unit increase in body mass index decreased LDL levels by `r round(effects["bmi"], 2)`%, while holding age and statin use constant. Finally, the largest effect was that of of statin use; subjects on statin had LDL levels `r round(effects["statin"], 2)`% lower than those subjects not on statin, while controlling for age and body mass index.

**Conclusion.** While controlling for age and body mass index, these results demonstrate a significant effect of statin use on lowering LDL cholesterol levels.


### 5. Now suppose interest focuses on whether BMI modifies the association between ln(LDL) and statin use.
#### 5.1 Extend the model in question 3 to address this question, and report the results. Assess whether the data provide strong evidence that BMI modifies the ln(LDL): statin use association.

Here we extend the multiple linear regression model to include a term modeling the interaction between body mass index and statin use:

The multivariate regression model is: 

$$
ln(LDL) = \beta_{1}age + \beta_{2}bmi + \beta_{3}statin + \beta_{4}(statin \enspace x \enspace bmi)
$$

```{r echo = FALSE}
# Run multivariate analysis, including all three predictor variables and an interaction between statin & BMI
interact.model <- lm(ln.LDL ~ age + bmi + statin + bmi * statin, data = hers.data)
interact.summary <- summary(interact.model)
interact.coeffs <- interact.summary$coefficients

# Deal with p-values as above (correcting for rounding down to zero)
interact.p.values <- interact.coeffs[, 4]
interact.p.values <- round(interact.p.values, 3)
interact.p.values[interact.p.values == 0] <- NA
interact.p.values <- as.character(interact.p.values)
interact.p.values[is.na(interact.p.values)] <- "< 0.001"
names(interact.p.values) <- rownames(interact.coeffs)
```

And the results of the linear model are:

|Parameter|Parameter Estimate|Standard Error|Pr($\neq0$)|
|---------|:----------------:|:------------:|:---------:|
| Intercept | `r round(interact.coeffs["(Intercept)", 1], 3)` | `r round(interact.coeffs["(Intercept)", 2], 3)` | `r interact.p.values["(Intercept)"]` |
| Age | `r round(interact.coeffs["age", 1], 3)` | `r round(interact.coeffs["age", 2], 3)` | `r interact.p.values["age"]` |
| BMI | `r round(interact.coeffs["bmi", 1], 3)` | `r round(interact.coeffs["bmi", 2], 3)` | `r interact.p.values["bmi"]` |
| Statin use | `r round(interact.coeffs["statin", 1], 3)` | `r round(interact.coeffs["statin", 2], 3)` | `r interact.p.values["statin"]` |
| Statin x BMI | `r round(interact.coeffs["bmi:statin", 1], 3)` | `r round(interact.coeffs["bmi:statin", 2], 3)` | `r interact.p.values["bmi:statin"]` |

#### 5.2 Explain why the main effect of statin use does not represent a meaningful effect in this model. 

When considering a model including the interaction between body mass index and statin, the effect of statin is non-significant. However, interpretation of this effect is not very useful, as it requires holding constant body mass index _and_ statin, as those are the two components of the interaction term. Interpreting the effect of statin on LDL requires including the interaction between statin and body mass index.

#### 5.3 Construct a new variable (centered bmi) defined as [bmi – mean(bmi)], where this mean is taken over the entire dataset. Re-fit the model in this question replacing bmi with this centered version, and report the results. Explain why the main effect of statin use now represents a meaningful quantity.

```{r echo = FALSE}
# Create new column of centered bmi
hers.data$centered.bmi <- hers.data$bmi - mean(hers.data$bmi)

# Run multivariate analysis, including all three predictor variables and an interaction between statin & BMI
centered.model <- lm(ln.LDL ~ age + centered.bmi + statin + centered.bmi * statin, data = hers.data)
centered.summary <- summary(centered.model)
centered.coeffs <- centered.summary$coefficients

# Deal with p-values as above (correcting for rounding down to zero)
centered.p.values <- centered.coeffs[, 4]
centered.p.values <- round(centered.p.values, 3)
centered.p.values[centered.p.values == 0] <- NA
centered.p.values <- as.character(centered.p.values)
centered.p.values[is.na(centered.p.values)] <- "< 0.001"
names(centered.p.values) <- rownames(centered.coeffs)
```

|Parameter|Parameter Estimate|Standard Error|Pr($\neq0$)|
|---------|:----------------:|:------------:|:---------:|
| Intercept | `r round(centered.coeffs["(Intercept)", 1], 3)` | `r round(centered.coeffs["(Intercept)", 2], 3)` | `r centered.p.values["(Intercept)"]` |
| Age | `r round(centered.coeffs["age", 1], 3)` | `r round(centered.coeffs["age", 2], 3)` | `r centered.p.values["age"]` |
| BMI | `r round(centered.coeffs["centered.bmi", 1], 3)` | `r round(centered.coeffs["centered.bmi", 2], 3)` | `r centered.p.values["centered.bmi"]` |
| Statin use | `r round(centered.coeffs["statin", 1], 3)` | `r round(centered.coeffs["statin", 2], 3)` | `r centered.p.values["statin"]` |
| Statin x BMI | `r round(centered.coeffs["centered.bmi:statin", 1], 3)` | `r round(centered.coeffs["centered.bmi:statin", 2], 3)` | `r centered.p.values["centered.bmi:statin"]` |


```{r echo = FALSE}
# Calculate the effect of statin on LDL (instead of ln(LDL))
statin.exp <- exp(x = centered.coeffs["statin", 1])
statin.effect <- abs(1 - statin.exp) * 100
```

When body mass index is transformed by the difference from the mean BMI, we can more easily interpret the effect of statin on the level of LDL cholesterol. Because the expected value of centered body mass index is zero, we only need consider the parameter estimate for statin when evaluating the effect of statin use on cholesterol (as opposed to the model in 5.1, where we needed to consider statin _and_ the BMI x statin interaction). Here we see a result similar to the univariate analysis of the effect of statin use on LDL cholesterol levels. Use of statin results in `r round(statin.effect, 2)`% decrease of LDL cholesterol (p `r centered.p.values["statin"]`).

***
***
The R code used in this assignment can be found in a corresponding R Markdown document at [https://github.com/jcoliver/bios-684/blob/master/homework-1-oliver.Rmd](https://github.com/jcoliver/bios-684/blob/master/homework-1-oliver.Rmd).
